---
title: "rr_proj1"
author: "Anand Singh"
date: "September 16, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

Unzip and read the csv data file
```{r}
unzip("activity.zip",exdir = "data")
activity <- read.csv("data/activity.csv", stringsAsFactors=FALSE)
head(activity)
```

Show summary statistics
```{r}
summary(activity)
str(activity)
activity$date<- as.Date(activity$date)
```


## What is mean total number of steps taken per day?
Calculate the total number of steps taken per day after removing NAs

```{r}
library(plyr)
library(dplyr)
totalsteps <- aggregate(steps ~ date, activity, sum, na.rm = TRUE)
totalsteps
```
Plot Histogram to show mean activity per day
```{r}
hist(totalsteps$steps, col = "blue", main = "Mean Activity per day")
```

Calculate and report the mean and median of the total number of steps taken per day
```{r}
mean(totalsteps$steps)
median(totalsteps$steps)
```

## What is the average daily activity pattern?
Aggregate steps by interval
```{r}
interval <- aggregate(steps ~ interval, activity, mean, na.rm = TRUE)
head(interval)
```

Plot Histogram to show average daily steps
```{r}
plot(interval$steps~ interval$interval, type = "l", col = "blue", main = "Average Daily Steps")
```

## Imputing missing values
Find out total number and percentage of missing values
```{r}
num_na <- sum(is.na(activity$steps))
num_na
pct_na <- round(mean(is.na(activity$steps))*100,1)
pct_na
```

Substitute missing values with average interval values
```{r}
activity2<- activity
nas<- is.na(activity2$steps)
avg_interval<- tapply(activity2$steps, activity2$interval, mean, na.rm=TRUE, simplify = TRUE)
activity2$steps[nas] <- avg_interval[as.character(activity2$interval[nas])]
sum(is.na(activity2))
head(activity2)
```
Plot Histogram to show all data including missing values
```{r}
totalsteps2 <- aggregate(steps ~ date, activity2, sum, na.rm = TRUE)
totalsteps2
hist(totalsteps2$steps, col = "blue", main = "Daily Steps including Missing Values")
```

Calculate and report the mean and median of the total number of steps taken per day again
```{r}
mean(totalsteps2$steps)
median(totalsteps2$steps)
```

## Are there differences in activity patterns between weekdays and weekends?

Introduce new column to categorize the type of day (weekday or weekend). Summarize steps based on interval and day type
```{r}
activity2<- activity2%>%
        mutate(daytype= ifelse(weekdays(activity$date)=="Saturday" | weekdays(activity2$date)=="Sunday", "Weekend", "Weekday"))
#avg_interval2<- tapply(activity2$steps, activity2$interval+activity2$weektype, mean, na.rm=TRUE, simplify = TRUE)
interval2 <- ddply(activity2, .(interval, daytype), summarize, Avg = mean(steps))
```
Plot panel plot to show difference between weekday and weekend
```{r}
library(lattice)
xyplot(Avg~interval|daytype, data=interval2, type="l",  layout = c(1,2),
       main="Average Steps per Interval Based on Type of Day", 
       ylab="Average Number of Steps", xlab="Interval")
```

This tells us that the activity pattern for weekdays and weekends are different